Arrows for Parallel Computations

### 2017-10-24 paper/content/basicSkeletons.tex
spellchecked

\subsection{Extending the Interface}
\label{sec:extending-interface}
With the |ArrowParallel| typeclass in place and implemented, we can now implement some further basic parallel interface functions. These are algorithmic skeletons that, however, mostly serve as a foundation to further, more specific algorithmic skeletons.

@@

\subsubsection{Heterogenous tasks}
\begin{figure}[tb]
	\includegraphics[scale=0.7]{images/parEval2}
	\caption{Schematic depiction of |parEval2|.}
	\label{fig:parEval2Img}
\end{figure}
We have only talked about the paralellization arrows of the same type until now. But sometimes we want to paralellize heterogenous types as well. However, we can implement such a |parEval2| combinator (Figs.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows |arr a b| and |arr c d| into a new parallel arrow |arr (a, c) (b, d)| quite easily with the help of the |ArrowChoice| typeclass. The idea is to use the |+++| combinator which combines two arrows |arr a b| and |arr c d| and transforms them into |arr (Either a c) (Either b d)| to get a common arrow type that we can then feed into |parEvalN|.

>>>>

\subsection{Extending the Interface}
\label{sec:extending-interface}
With the |ArrowParallel| type class in place and implemented, we can now implement some further basic parallel interface functions. These are algorithmic skeletons that, however, mostly serve as a foundation to further, more specific algorithmic skeletons.

@@

\subsubsection{Heterogeneous tasks}
\begin{figure}[tb]
	\includegraphics[scale=0.7]{images/parEval2}
	\caption{Schematic depiction of |parEval2|.}
	\label{fig:parEval2Img}
\end{figure}
We have only talked about the parallelization arrows of the same type until now. But sometimes we want to parallelize heterogeneous types as well. However, we can implement such a |parEval2| combinator (Figs.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows |arr a b| and |arr c d| into a new parallel arrow |arr (a, c) (b, d)| quite easily with the help of the |ArrowChoice| type class. The idea is to use the |+++| combinator which combines two arrows |arr a b| and |arr c d| and transforms them into |arr (Either a c) (Either b d)| to get a common arrow type that we can then feed into |parEvalN|.

### 2017-10-24 paper/content/benchmarks.tex
spellchecked

We have tested our parallel DSL and algorithmic skeletons implemented
in it. Benchmarks were conducted both in a shared and in a distributed
memory setting. All benchmarks were done on the ``Glasgow grid'', consisting of
16 machines with 2 Intel\SymbReg~Xeon\SymbReg~E5-2640 v2 and 64 GB of DDR3 RAM each. Each processor has 8 cores and 16 (hyperthreaded) threads with a base frequency of 2 GHz and a turbo frequency of 2.50 GHz. This results in a total of 256 cores and 512 threads for the whole cluster. The operating system was Ubuntu 14.04 LTS with Kernel 3.19.0-33. Non-surprisingly, we found that hyperthreaded 32 cores do not behave in the same manner as real 16 cores (numbers here for a single machine). We disregarded the hyperthreading ability in most of the
cases.

@@

\subsection{Effect of hyperthreading}

The PArrows version of \rmtest on a single node of the Glasgow grid
showed almost linear speedup (Fig.~\ref{fig:bench-rm-sm}). The speedup
of 64-task PArrows/Eden at 16 real cores version was 13.65, the
efficiency was 85.3\%. However, if we increase the number of
requested cores to be 32 -- \ie if we use hyperthreading on 16 real
cores -- the speedup does not increase that well. It is merely 15.99
for 32 tasks with PArrows/Eden. It is worse for other backends.  As
for 64 tasks, we obtain the speedup of 16.12 with PArrows/Eden at 32
hyperthreaded cores and only 13.55 with PArrows/GpH. Efficiency is 50.4\% and 42.3\%, respectively. The Eden
version used here was Eden~CP, the \enquote{share nothing} SMP build.

In the distributed memory setting the same effect ensues. We obtain
plummeting speedup of 124.31 at 512 hyperthreaded cores, whereas it was
213.172 for 256 real cores. Apparently, hyperthreading in the Glasgow
grid fails to execute two parallel Haskell processes with full-fledged
parallelism. For this reason, we did not regard hyperthreaded cores in
our speedup plots in Figs.~\ref{fig:bench-rm-sm}--\ref{fig:sudokuSMBenchmark}.

@@

To show that PArrows induce very small overhead in a distributed context as well, we compare the original Eden
versions of the benchmark to its PArrows-enabled counterpart in the \rmtest, \torustest and \jacobitest benchmarks. We plot execution time differences between measurements for
PArrows and the corresponding backend in a separate plot
(Figs.~\ref{fig:bench-rm-dist}--\ref{fig:torusBenchmark}). As an example, the differences range in
about 0.5\seconds for the execution time of 46\seconds on 256 cores
for distributed \rmtest with PArrows and Eden. For these comparisons, the plots show absolute
time differences that are not relative \wrt the total execution time.
Furthermore, the error bars ends were computed from pointwise maximum of both standard
deviations from both measurements for PArrows and non-PArrows
versions. These are the values provided by the |bench| package that we
used for benchmarking. We call a difference between two versions
significant when the border of the error bar of absolute time
difference is above or below zero. In other words: the time
difference is significant if it is above measurement error.

@@

As the last benchmark in this paper we present the \sudokutest in Fig.~\ref{fig:sudokuSMBenchmark} running in a shared memory setting. Here we see all three SM backends performing similarly again like in the \rmtest SM benchmarks in Figs.~\ref{fig:sudokuSMBenchmark} and \ref{fig:sudokuSMBenchmark16000}. However, we notice that the GpH backend seems to choke on a bigger input (Fig.~\ref{fig:sudokuSMBenchmark16000}). This is due to the benchmark only using |parMap| instead of a chunking variant -- however we did not change that for simplicity's sake. This issue is reflected by debug output which shows that of 16000 sparks being created (one for each sudoku) only 8365 were converted (executed) with the rest (7635) overflowing the runtime spark pool. Another remarkable finding is that the Eden backend seems to lag behind for $\leq$16 threads, but manages to pull ahead noticeably with all 32 threads of the system in use.

>>>>

We have tested our parallel DSL and algorithmic skeletons implemented
in it. Benchmarks were conducted both in a shared and in a distributed
memory setting. All benchmarks were done on the ``Glasgow grid'', consisting of
16 machines with 2 Intel\SymbReg~Xeon\SymbReg~E5-2640 v2 and 64 GB of DDR3 RAM each. Each processor has 8 cores and 16 (hyper-threaded) threads with a base frequency of 2 GHz and a turbo frequency of 2.50 GHz. This results in a total of 256 cores and 512 threads for the whole cluster. The operating system was Ubuntu 14.04 LTS with Kernel 3.19.0-33. Non-surprisingly, we found that hyper-threaded 32 cores do not behave in the same manner as real 16 cores (numbers here for a single machine). We disregarded the hyper-threading ability in most of the
cases.

@@

\subsection{Effect of hyper-threading}

The PArrows version of \rmtest on a single node of the Glasgow grid
showed almost linear speedup (Fig.~\ref{fig:bench-rm-sm}). The speedup
of 64-task PArrows/Eden at 16 real cores version was 13.65, the
efficiency was 85.3\%. However, if we increase the number of
requested cores to be 32 -- \ie if we use hyper-threading on 16 real
cores -- the speedup does not increase that well. It is merely 15.99
for 32 tasks with PArrows/Eden. It is worse for other backends.  As
for 64 tasks, we obtain the speedup of 16.12 with PArrows/Eden at 32
hyper-threaded cores and only 13.55 with PArrows/GpH. Efficiency is 50.4\% and 42.3\%, respectively. The Eden
version used here was Eden~CP, the \enquote{share nothing} SMP build.

In the distributed memory setting the same effect ensues. We obtain
plummeting speedup of 124.31 at 512 hyper-threaded cores, whereas it was
213.172 for 256 real cores. Apparently, hyper-threading in the Glasgow
grid fails to execute two parallel Haskell processes with full-fledged
parallelism. For this reason, we did not regard hyper-threaded cores in
our speedup plots in Figs.~\ref{fig:bench-rm-sm}--\ref{fig:sudokuSMBenchmark}.

@@

To show that PArrows induce very small overhead in a distributed context as well, we compare the original Eden
versions of the benchmark to its PArrows-enabled counterpart in the \rmtest, \torustest and \jacobitest benchmarks. We plot execution time differences between measurements for
PArrows and the corresponding backend in a separate plot
(Figs.~\ref{fig:bench-rm-dist}--\ref{fig:torusBenchmark}). As an example, the differences range in
about 0.5\seconds for the execution time of 46\seconds on 256 cores
for distributed \rmtest with PArrows and Eden. For these comparisons, the plots show absolute
time differences that are not relative \wrt the total execution time.
Furthermore, the error bars ends were computed from point-wise maximum of both standard
deviations from both measurements for PArrows and non-PArrows
versions. These are the values provided by the |bench| package that we
used for bench-marking. We call a difference between two versions
significant when the border of the error bar of absolute time
difference is above or below zero. In other words: the time
difference is significant if it is above measurement error.

@@

As the last benchmark in this paper we present the \sudokutest in Fig.~\ref{fig:sudokuSMBenchmark} running in a shared memory setting. Here we see all three SM backends performing similarly again like in the \rmtest SM benchmarks in Figs.~\ref{fig:sudokuSMBenchmark} and \ref{fig:sudokuSMBenchmark16000}. However, we notice that the GpH backend seems to choke on a bigger input (Fig.~\ref{fig:sudokuSMBenchmark16000}). This is due to the benchmark only using |parMap| instead of a chunking variant -- however we did not change that for simplicity's sake. This issue is reflected by debug output which shows that of 16000 sparks being created (one for each Sudoku) only 8365 were converted (executed) with the rest (7635) overflowing the runtime spark pool. Another remarkable finding is that the Eden backend seems to lag behind for $\leq$16 threads, but manages to pull ahead noticeably with all 32 threads of the system in use.

### 2017-10-24 paper/content/futures.tex
spellchecked

This is in fact only a problem in distributed memory (in the scope of this paper) and we should  allow nodes to communicate directly with each other. Eden already ships with \enquote{remote data} that enable this \cite{AlGo03a,Dieterle2010}.
But as we want code with our DSL to be implementation agnostic, we have to wrap this context. We do this with the |Future| typeclass (Fig.~\ref{fig:future}).
\begin{figure}[h]
\begin{code}
class Future fut a | a -> fut where
    put :: (Arrow arr) => arr a (fut a)
    get :: (Arrow arr) => arr (fut a) a
\end{code}
\caption{Definition of the |Future| typeclass.}
\label{fig:future}
\end{figure}
Since |RD| is only a type synonym for a communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as Fig.~\ref{fig:RDFuture} shows. %This is due to the same reason we had to introduce a wrapper for |Strategy a| in the GpH Haskell implementation of |ArrowParallel| in Section~\ref{sec:parrows:multicore}.
Technical details are in Appendix, in Section~\ref{app:omitted}.

For the |Par| Monad and GpH Haskell backends, we can simply use |BasicFuture|s (Fig.~\ref{fig:BasicFuture}), which are just simple wrappers around the actual data with boiler-plate logic so that the typeclass is satisfied. This is because the concept of a |Future| does not change anything for shared-memory execution as there are no communication problems to fix. Nevertheless, we require a common interface so the parallel Arrows are portable across backends.
\begin{figure}[tb]
\begin{code}
data BasicFuture a = BF a

>>>>

This is in fact only a problem in distributed memory (in the scope of this paper) and we should  allow nodes to communicate directly with each other. Eden already ships with \enquote{remote data} that enable this \cite{AlGo03a,Dieterle2010}.
But as we want code with our DSL to be implementation agnostic, we have to wrap this context. We do this with the |Future| type class (Fig.~\ref{fig:future}).
\begin{figure}[h]
\begin{code}
class Future fut a | a -> fut where
    put :: (Arrow arr) => arr a (fut a)
    get :: (Arrow arr) => arr (fut a) a
\end{code}
\caption{Definition of the |Future| type class.}
\label{fig:future}
\end{figure}
Since |RD| is only a type synonym for a communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as Fig.~\ref{fig:RDFuture} shows. %This is due to the same reason we had to introduce a wrapper for |Strategy a| in the GpH Haskell implementation of |ArrowParallel| in Section~\ref{sec:parrows:multicore}.
Technical details are in Appendix, in Section~\ref{app:omitted}.

For the |Par| Monad and GpH Haskell backends, we can simply use |BasicFuture|s (Fig.~\ref{fig:BasicFuture}), which are just simple wrappers around the actual data with boiler-plate logic so that the type class is satisfied. This is because the concept of a |Future| does not change anything for shared-memory execution as there are no communication problems to fix. Nevertheless, we require a common interface so the parallel Arrows are portable across backends.
\begin{figure}[tb]
\begin{code}
data BasicFuture a = BF a

### 2017-10-24 paper/content/motivation.tex
wip
spellchecked

%Our Arrows-based interface is a high-level one.
We provide an Arrows-based typeclass and implementations for the three above mentioned parallel Haskells.
Instead of 
introducing a new low-level parallel backend in order to implement our
Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
is defined as a common interface with varying implementations in
the existing parallel Haskells. Thus, we not only define a parallel programming interface in a
novel manner - we tame the zoo of parallel Haskells. We provide a
common \ptcomment{quantify, e.g. less than 10 \%}, very low-penalty programming interface that allows to switch
the parallel backends at will. Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.

@@

We do not reimplement all the parallel internals, as we host this functionality in the |ArrowParallel| typeclass, which abstracts all parallel implementation logic. This way, we can have multiple backends - currently a GpH, a |Par| Monad and an Eden version are supported. The backends can easily be swapped, so we are not bound to any specific one. So as an example, during development, we can run the program in a simple GHC-compiled variant using a GpH backend and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the |ArrowParallel| instance and compiling with Eden's GHC variant.

We introduce a |Future| typeclass in Section \ref{sec:futures} to enable direct communication of data between nodes in a distributed memory setting similar to Eden's Remote Data (|RD|). Direct communication is useful in a distributed memory setting because they allow for inter-node communication without blocking the master-node. As the |Par| Monad and GpH backends (in their current form) only support shared memory execution and do not have to fix any communication problems, we only use a wrapping dummy |BasicFuture|.

>>>>

%Our Arrows-based interface is a high-level one.
We provide an Arrows-based type class and implementations for the three above mentioned parallel Haskells.
Instead of 
introducing a new low-level parallel backend in order to implement our
Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
is defined as a common interface with varying implementations in
the existing parallel Haskells. Thus, we not only define a parallel programming interface in a
novel manner - we tame the zoo of parallel Haskells. We provide a
common \ptcomment{quantify, e.g. less than 10 \%}, very low-penalty programming interface that allows to switch
the parallel backends at will. Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.

@@

We do not reimplement all the parallel internals, as we host this functionality in the |ArrowParallel| type class, which abstracts all parallel implementation logic. This way, we can have multiple backends -- currently a GpH, a |Par| Monad and an Eden version are supported. The backends can easily be swapped, so we are not bound to any specific one. So as an example, during development, we can run the program in a simple GHC-compiled variant using a GpH backend and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the |ArrowParallel| instance and compiling with Eden's GHC variant.

We introduce a |Future| type class in Section \ref{sec:futures} to enable direct communication of data between nodes in a distributed memory setting similar to Eden's Remote Data (|RD|). Direct communication is useful in a distributed memory setting because they allow for inter-node communication without blocking the master-node. As the |Par| Monad and GpH backends (in their current form) only support shared memory execution and do not have to fix any communication problems, we only use a wrapping dummy |BasicFuture|.

### 2017-10-24 paper/content/topologySkeletons.tex
spellchecked

If we used the original definition of |parEvalN|, however, these skeletons would produce an infinite loop with the GpH and Par Monad backends which during runtime would result in the program crashing. This materializes with the usage of |loop| of the |ArrowLoop| typeclass and is probably due to the way their respective parallelism engines work internally. \mbcomment{okay so?} As these skeletons probably do not make any practical sense besides for testing with these backends anyways (because of the shared memory between the threads), we create an extra abstraction layer for the original |parEvalN| in these skeletons called |evalN| in the |FutureEval| typeclass. This allows us for selective enabling and disabling of parallelism.
\begin{code}
class ArrowParallel arr a b conf => FutureEval arr a b conf where
    evalN :: (ArrowParallel arr a b conf) => conf -> [arr a b] -> arr [a] [b]
\end{code}
As Eden has no problems with the looping skeletons, we declare a delegating instance:
\begin{code}
instance ArrowParallel arr a b conf => FutureEval arr a b conf where
    evalN = parEvalN
\end{code}
The Par Monad and GpH backends have parallelism disabled in their instance of |FutureEval|. This way the skeletons can still run without errors on shared-memory machines and still be used to test programs locally.
\begin{code}
instance (Arrow arr, ArrowChoice arr, ArrowApply arr,
    ArrowParallel arr a b conf) => FutureEval arr a b conf where
    evalN _ = listApp
\end{code}

@@

The parallel |pipe| skeleton is semantically equivalent to folding over a list |[arr a a]| of arrows with |>>>|, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine. We implement this skeleton using the |ArrowLoop| typeclass which gives us the |loop :: arr (a, b) (c, b) -> arr a c| combinator which allows us to express recursive fix-point computations in which output values are fed back as input. For example %this
\begin{code}
loop (arr (\(a, b) -> (b, a:b)))
\end{code}
which is the same as
\begin{code}
loop (arr snd &&& arr (uncurry (:)))
\end{code}
defines an arrow that takes its input |a| and converts it into an infinite stream |[a]| of it. Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call |evalN conf fs| inside the second argument of |&&&| and then only picking the first element of the resulting list with |arr last|.

@@

Another version of |>>>| is:
\begin{code}
f parcomp g = (f . put) >>> (get . g)
\end{code}
It does not launch both arrows |f| and |g| in parallel, but allows for more smooth data commuinication between them. Basically, it is a |Future|-lifted \emph{sequential} |>>>|, a way to compose parallel Arrows efficiently.

@@

\subsection{Torus skeleton}\label{sec:torus}
\begin{figure}
	\includegraphics[scale=0.75]{images/torus}
	\caption{Schematic depiction of the |torus| skeleton.}
	\label{fig:ringTorusImg}
\end{figure}
If we take the concept of a |ring| from Section~\ref{sec:ring} one dimension further, we obtain a |torus| skeleton (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}). Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
With our Parallel Arrows we re-implement the |torus| combinator\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}.} from Eden---yet again with the help of the |ArrowLoop| typeclass.

@@

This action is, however,  more complicated than in the ring case as we have one more dimension of inputs to be transformed. We first have to |shuffle| all the inputs to then pass it into |evalN conf (repeat (ptorus f))| which yields |[(d, fut a, fut b)]|. We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with |arr length >>> arr unshuffle| with the use of |app :: arr (arr a b, a) c| from the |ArrowApply| typeclass. Finally, we unpack this matrix |[[[(d, fut a, fut b)]]| with |arr (map unzip3) >>> arr unzip3 >>> threetotwo| to get  |([[d]], ([[fut a]], [[fut b]]))|.

>>>>

If we used the original definition of |parEvalN|, however, these skeletons would produce an infinite loop with the GpH and Par Monad backends which during runtime would result in the program crashing. This materializes with the usage of |loop| of the |ArrowLoop| type class and is probably due to the way their respective parallelism engines work internally. \mbcomment{okay so?} As these skeletons probably do not make any practical sense besides for testing with these backends anyways (because of the shared memory between the threads), we create an extra abstraction layer for the original |parEvalN| in these skeletons called |evalN| in the |FutureEval| type class. This allows us for selective enabling and disabling of parallelism.
\begin{code}
class ArrowParallel arr a b conf => FutureEval arr a b conf where
    evalN :: (ArrowParallel arr a b conf) => conf -> [arr a b] -> arr [a] [b]
\end{code}
As Eden has no problems with the looping skeletons, we declare a delegating instance:
\begin{code}
instance ArrowParallel arr a b conf => FutureEval arr a b conf where
    evalN = parEvalN
\end{code}
The Par Monad and GpH backends have parallelism disabled in their instance of |FutureEval|. This way the skeletons can still run without errors on shared-memory machines and still be used to test programs locally.
\begin{code}
instance (Arrow arr, ArrowChoice arr, ArrowApply arr,
    ArrowParallel arr a b conf) => FutureEval arr a b conf where
    evalN _ = listApp
\end{code}

@@

The parallel |pipe| skeleton is semantically equivalent to folding over a list |[arr a a]| of arrows with |>>>|, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine. We implement this skeleton using the |ArrowLoop| type class which gives us the |loop :: arr (a, b) (c, b) -> arr a c| combinator which allows us to express recursive fix-point computations in which output values are fed back as input. For example %this
\begin{code}
loop (arr (\(a, b) -> (b, a:b)))
\end{code}
which is the same as
\begin{code}
loop (arr snd &&& arr (uncurry (:)))
\end{code}
defines an arrow that takes its input |a| and converts it into an infinite stream |[a]| of it. Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call |evalN conf fs| inside the second argument of |&&&| and then only picking the first element of the resulting list with |arr last|.

@@

Another version of |>>>| is:
\begin{code}
f parcomp g = (f . put) >>> (get . g)
\end{code}
It does not launch both arrows |f| and |g| in parallel, but allows for more smooth data communication between them. Basically, it is a |Future|-lifted \emph{sequential} |>>>|, a way to compose parallel Arrows efficiently.

@@

\subsection{Torus skeleton}\label{sec:torus}
\begin{figure}
	\includegraphics[scale=0.75]{images/torus}
	\caption{Schematic depiction of the |torus| skeleton.}
	\label{fig:ringTorusImg}
\end{figure}
If we take the concept of a |ring| from Section~\ref{sec:ring} one dimension further, we obtain a |torus| skeleton (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}). Every node sends and receives data from horizontal and vertical neighbours in each communication round.
With our Parallel Arrows we re-implement the |torus| combinator\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}.} from Eden---yet again with the help of the |ArrowLoop| type class.

@@

This action is, however,  more complicated than in the ring case as we have one more dimension of inputs to be transformed. We first have to |shuffle| all the inputs to then pass it into |evalN conf (repeat (ptorus f))| which yields |[(d, fut a, fut b)]|. We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with |arr length >>> arr unshuffle| with the use of |app :: arr (arr a b, a) c| from the |ArrowApply| type class. Finally, we unpack this matrix |[[[(d, fut a, fut b)]]| with |arr (map unzip3) >>> arr unzip3 >>> threetotwo| to get  |([[d]], ([[fut a]], [[fut b]]))|.
